On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-09-13 manuscript/pnas_revision.tex
more revisions

\section{Visualizing the posterior mean random effects}

Figure

\begin{figure}[t]
	\centering
	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.  Notably, we infer a roughly 1.8-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.  Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}\label{fig:randEffs}
\end{figure}

>>>>

\section{Visualizing the posterior mean random effects}\label{sec:randomEffs}

Figure \ref{fig:randEffs} displays posterior means of the exponentiated random effects, which one may interpret as multiplicative deviations from the fixed effects' contributions to the generator matrix.  Whereas the vast majority of exponentiated random effects have posterior means close to 1, indicating no deviation from the fixed-effect model, a few exhibit posterior means that are significantly greater than 1.  In particular, the rate element corresponding to transfer from the US to Hubei, CN, exhibits a 1.81-fold random-effect derived increase to the fixed-effect component.  This large multiplicative increase agrees with, but goes beyond, the influence that the Hubei asymmetry holds for the entire generator matrix model.

\begin{figure}[t]
	\centering
	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.  Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.  Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}\label{fig:randEffs}
\end{figure}

### 2023-09-13 manuscript/revision_letter.tex
more revisions

\begin{reply}
This is what 
\end{reply}

\begin{enumerate}	
\setcounter{enumi}{5}	
	\item The ranges for predictor matrices and posterior mean rate matrix are different. However, the color palettes are identical, which makes it extremely hard to compare the impact of each predictor. For example, I cannot see how the sentence “... but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.” makes sense visually from Fig. 2 and 3. It also doesn’t make sense to me conceptually, because air traffic proximity is the only significant predictor.
\end{enumerate}

>>>>

\setcounter{figure}{4}
\begin{figure}[t]
	\centering
	\includegraphics[width=0.8\linewidth]{figures/randEffects.pdf}
	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.  Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.  Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}%\label{fig:randEffs}
\end{figure}

\begin{reply}
This is a good question.  Figure \ref{fig:randEffs} show posterior means of the exponentiated random effects, which amount to multiplicative deviations from the fixed-effect model.  As one might expect, only a small number of random effects deviate from 0, but those that do deviate are highly suggestive.  We stop short of interpreting exactly why the transition from the US to Hubei, CN, exhibits largest deviation (a 1.81-fold increase).  We have added this figure and the following short discussion to the appendix:
	\begin{quote}
		Figure \ref{fig:randEffs} displays posterior means of the exponentiated random effects, which one may interpret as multiplicative deviations from the fixed effects' contributions to the generator matrix.  Whereas the vast majority of exponentiated random effects have posterior means close to 1, indicating no deviation from the fixed-effect model, a few exhibit posterior means that are significantly greater than 1.  In particular, the rate element corresponding to transfer from the US to Hubei, CN, exhibits a 1.81-fold random-effect derived increase to the fixed-effect component.  This large multiplicative increase agrees with, but goes beyond, the influence that the Hubei asymmetry holds for the entire generator matrix model. 
	\end{quote}	
\end{reply}

\begin{enumerate}	
\setcounter{enumi}{5}	
	\item The ranges for predictor matrices and posterior mean rate matrix are different. However, the color palettes are identical, which makes it extremely hard to compare the impact of each predictor. For example, I cannot see how the sentence ``... but one may also see the influence of the Hubei asymmetry in, e.g., the squares corresponding to travel between Guangdong and Hubei provinces.'' makes sense visually from Fig. 2 and 3. It also doesn’t make sense to me conceptually, because air traffic proximity is the only significant predictor.
\end{enumerate}

\begin{reply}
The ranges are different because the predictors are defined on the log scale, whereas the generator matrix is obtained by the (element-wise) exponentiation of the predictors times their respective regression coefficients plus the random-effects matrix.  Even though the Hubei asymmetry is not statistically significant (i.e., its 95\% credible interval contains 0), it still has a non-negligible effect size ($\approx 0.25$ posterior mean per Fig. 3a) that helps explain its appearance in Figure 3b.
\end{reply}
